QCL: Added value of differential correlation

### 2011-11-24 inst/doc/article.Rnw
Uploaded the tex file by accident

<<echo=FALSE,results=hide>>=
  require(qcl)
  set.seed(61777369)
  data(ath.metabolites) # Arabidopsis Thaliana dataset
  qcl_result <- QCLscan(ath.metab$genotypes, ath.metab$phenotypes)
@

@@

\emph{\sffamily Example 1. A. Thaliana}

QCL mapping was performed on 24 characterized metabolities measured in Arabidopsis 
Thaliana. A significance threshold for QCL mapping was determined by using 1000 
permutations, the outcomes can be seen in table XX. We opted for the stringent 
threshold at P < 0.001 to use as 'significant' QCL score for further analysis.

Overview plots of the results of the QCL mapping in Arabidopsis can be found in 
Figure XX Phenotype to Marker interaction, which is comparable to the QTL mapping 
(Fig. XX). The additional information obtained by QCL mapping is shown in a heatmap 
of the Phenotype to Phenotype interaction matrix (figure XX) and the same information 
but now represented as an interaction network is seen in figure XX
\begin{table}[ht]
\begin{tabular}{ l | c c }
P-value & QCL threshold Arabidopsis & QCL threshold Saccharomyces\\
\hline
5 \% & 0.220 & 0.436\\
1 \% & 0.294 & 0.491\\
0.1 \% & 0.375 & 0.556\\
\end{tabular}
\caption{QCL thresholds obtained after 1000 permutations}
\label{permutations}
\end{table}

\emph{\sffamily Example 1. S. Cerevisiae}

QCL mapping was performed on

@@

\begin{figure}[ht]
  \centerline{\includegraphics{QCL.pdf}}
  \caption{QCL mapping: The underlying hypothesis of QCL mapping. Seen here is a network of 3 phenotypes, in 
  which expression of P1, P2 and P3 is measured, and we want to infer the hidden relation between P1 
  and P2. In our model a genetic pertubation is causing P1 to show a QTL at a certain marker. We expect variation 
  induced by P1 to transfer to P2, However (like in most biological systems) expression of P2 is tightly controlled, 
  and variation seen in phenotype 1 wil not induce a QTL in phenotype 2. However when we partition individuals based 
  on the genotype at the QTL marker of P1, we can make use of the correlation structure between P1 and P2. Allowing 
  us to infer the relationship between phenotype 1 and 2, when we see difference in correlation between P1A and P2A 
  and P1B and P2B. We don't observe this difference in correlation between P3 and P1, meaning there is no relation 
  between P1 and P3}
  \label{QCL_img}
\end{figure}

\begin{figure}
  \centering
<<singletrait, fig=TRUE, echo=FALSE, height=3.5>>=
plot(qcl_result, 2, do.legend=FALSE)
@ 
  \caption{Single trait QCL profile (A. Thaliana) shows the number of phenotypes that have a significant 
  QCL with the scanned phenotype at different QCL significance thresholds.\label{fig:singletrait}}
\end{figure}

\begin{figure}[ht]
  \centering
<<p2mmatrix, fig=TRUE, echo=FALSE, height=3.5>>=
par(mar=c(3,8,1.1,1.1))
image(qcl_result, 0.5, against="markers",do.grid=FALSE)
@ 
  \caption{Phenotypes x Markers matrix (A. Thaliana) this is 'comparable' to a QTL scan on multiple traits.\label{fig:p2mmatrix}}
\end{figure}

>>>>

<<echo=FALSE,results=hide>>=
  require(qcl)
  set.seed(200)
  data(ath.metabolites) # Arabidopsis Thaliana dataset
  cat("QCL scan\n")
  qcl_result <- QCLscan(ath.metab$genotypes, ath.metab$phenotypes)
  cat("QTL scan\n")
  qtl_result <- QTLscan(ath.metab$genotypes, ath.metab$phenotypes)
  cat("QCL permutation\n")
  qcl_perms <- QCLpermute(ath.metab$genotypes, ath.metab$phenotypes,1:24,20, n.cores = 3)
  #qcl_perms <- vector("list",24)
  #idx <- 1
  #for(p in 1:24){
  #  qcl_perms[[idx]] <- read.QCLpermute("permutations", p, 300,TRUE)
  #  idx <- idx+1
  #}
@

@@

  \subsection{Example 1. A. Thaliana}
  QCL mapping was performed on 24 characterized metabolities measured in Arabidopsis 
  Thaliana. A significance threshold for QCL mapping was determined for each trait 
  by using 1000 permutations, the outcomes can be seen in Fig \ref{fig:histPermAra}. 
  We used a very stringent QCL threshold at P < 0.001 to use as 'significant' QCL 
  score for further analysis.

  \emph{\sffamily Significance}\\
  The major observations is that we cannot use a 'flat' cut-off for all traits, each 
  trait shows a distinct NULL-distribution. A squared correlation difference of 0.5 is 
  significant for X3.Hydroxypropyl (green). But for X4.Methylsulfinylbutyl we need a 
  difference in squared correlation larger then 0.7 (yellow).

  \emph{\sffamily Single trait QCL}\\
  The 24 QCL profiles were compared to the QTL profiles previously determined. We observed 3 cases:\\
  1) Additional QCL found, where no QTL was present as seen in the QCL profile X3.Hydroxypropyl 
  (Fig. \ref{fig:singletraitLOD}), these are the cases in which the loss of variation in the genes 
  underneath the QCL is not causing variation differences in the observed phenotype, but still 
  we're able to detect the correlation differences.\\
  2) QCL co-localized with QTL, in this situation QCL gives a detailed view of which phenotypes 
  are underneath its peek, we interpret this as variation transfered from the phenotypes underneath 
  the QCL to the phenotype investigated.\\
  3) QTL with no QCL, we interpret this as the source of the variation. Analysis in this way of 
  all 24 Single trait QCL profiles combined with their QTL profiles allows us to reconstruct the 
  source of natural variation and follow it down the network.

  \emph{\sffamily QCL combined with QTL}\\
  Plots of the results of the QCL mapping in Arabidopsis can be found in 
  Figure XX Phenotype to Marker interaction, which is comparable to the QTL mapping 
  (Fig. XX). The additional information obtained by QCL mapping is shown in a heatmap 
  of the Phenotype to Phenotype interaction matrix (figure XX) and the same information 
  but now represented as an interaction network is seen in figure XX and in the QCL plot

  \emph{\sffamily Reconstructed network}\\
  Using the derived data we can begin to reconstruct the network. We'll start with a metabolite 
  which shows mappable variation changes, but no QCL colocatizing (Inducer).

  \subsection{Example 2. S. Cerevisiae}
  QCL mapping is computationaly expensive (We need 1000+ permutations for each phenotype), 
  and for the S. Cerevisiae we pre-selected a background of 300 probes which showed the 
  highest variation, and had been previously been identified as having QTLs. Significance 
  thresholds for QCL mapping were determined for each trait by using 1000 permutations 
  ({\emph Data not shown}).

  \emph{\sffamily Single trait QCL}\\

@@

\begin{figure}[ht]
  \centerline{\includegraphics{QCL.pdf}}
  \caption{QCL mapping: The underlying hypothesis of QCL mapping. Seen here is a network of 3 phenotypes, in 
  which expression of P1, P2 and P3 is measured, and we want to infer the hidden relation between P1 
  and P2. In our model a genetic pertubation is causing P1 to change in expression showing a QTL at a marker. We expect variation 
  induced by P1 to transfer to P2, However (like in most biological systems) expression of P2 is tightly controlled, 
  and variation seen in phenotype 1 wil not induce a QTL in phenotype 2. However when we partition individuals based 
  on the genotype at the QTL marker of P1, we can make use of the correlation structure between P1 and P2. Allowing 
  us to infer the relationship between phenotype 1 and 2, when we see difference in correlation between P1A and P2A 
  and P1B and P2B. We don't observe this difference in correlation between P3 and P1, meaning there is no relation 
  between P1 and P3}
  \label{QCL_img}
\end{figure}

\begin{figure}[ht]
  \centering
<<permutation_scores, fig=TRUE, echo=FALSE, height=3.5>>=
op <- par(cex=0.5)
plot(c(0,0.8),c(0,400),type='n',main="QCL scores permutations",ylab="Frequency",xlab="Difference in correlation^2")
hist(qcl_perms[[3]],add=T, col="yellow")
hist(qcl_perms[[2]],add=T, col="red")
hist(qcl_perms[[1]],add=T, col="green")
legend("topright",colnames(ath.metab$phenotypes)[1:3],col=c("yellow","red","green"),lwd=6)
op <- par(cex=1)
@ 
  \caption{QCL scores obtained during permutations (A. Thaliana) additional information obtained by QCL mapping, 
  which can be used to reconstruct the underlying genetic architecture.\label{fig:histPermAra}}
\end{figure}

\begin{figure}
  \centering
<<singletrait, fig=TRUE, echo=FALSE, height=3.0>>=
op <- par(cex=0.5)
plotAsLOD(qcl_result, qcl_perms, pheno.col=1)
op <- par(cex=1)
@
  \caption{QCL profile (A. Thaliana, X3.Hydroxypropyl) transformed to LOD scores compared to the QTL profile of the same phenotype, We see an 
  additional significant peek for this phenotype (Between Marker 50 to 60) \label{fig:singletrait}}
\end{figure}

\begin{figure}
  \centering
<<singletraitLOD, fig=TRUE, echo=FALSE, height=3.0>>=
op <- par(cex=0.5)
plotAsStackedHist(qcl_result, qcl_perms, pheno.col=1)
op <- par(cex=1)
@ 
  \caption{QCL profile (A. Thaliana, X3.Hydroxypropyl) transformed to LOD scores. Shown is the likelyhood contribution of each of 
  the 24 metabolites to the QCL profile. \label{fig:singletraitLOD}}
\end{figure}

\begin{figure}[ht]
  \centering
<<p2mmatrix, fig=TRUE, echo=FALSE, height=3.0>>=
par(mar=c(3,8,1.1,1.1))
image(qcl_result, qcl_perms, against="markers",do.grid=FALSE)
@ 
  \caption{Phenotypes x Markers matrix (A. Thaliana) this is 'comparable' to a QTL scan on multiple traits.\label{fig:p2mmatrix}}
\end{figure}

\begin{figure}[ht]
  \centering
<<QTLmatrix, fig=TRUE, echo=FALSE, height=3.0>>=
par(mar=c(3,8,1.1,1.1))
image(1:ncol(qtl_result),1:nrow(qtl_result),t(qtl_result),col=c("white",gray.colors(100)[100:1]), yaxt="n", xaxt="n", ylab="", xlab="")
axis(2,rownames(qtl_result),at=1:nrow(qtl_result),las=2,cex.axis=0.5)
axis(1,colnames(qtl_result),at=1:ncol(qtl_result),las=2,cex.axis=0.3)
@ 
  \caption{Heatmap showing QTL scores for all 24 metabolities in A. Thaliana.\label{fig:QTLmatrix}}
\end{figure}
